Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages

### 2011-04-06 talk/icooolps2011/paper.tex
fix some things

It has long been an objective of the partial evaluation community to
automatically produce compilers from interpreters. There has been a recent
renaissance of this idea around the approach of tracing just-in-time
compilers. A number of projects have attempted this approach. SPUR \cite{bebenita_spur:_2010} is
a tracing JIT for .NET together with a JavaScript implementation in C\#. PyPy
\cite{armin_rigo_pypys_2006} contains a tracing JIT for RPython \cite{davide_ancona_rpython:_2007} (a restricted
subset of Python). This JIT is then used to trace a number of languages
implementations written in RPython. A number of other experiments in this
directions were done, such as an interpreter for Lua in JavaScript, which is run
on and optimized with a tracing JIT for JavaScript
\cite{yermolovich_optimization_2009}.

These projects have in common that they work one meta-level down, providing a tracing JIT for the
language used to implement the dynamic language, and not for the dynamic language itself.
The tracing JIT then will trace through the object model of the dynamic
language implementation. This makes the object model transparent to the tracer
and its optimizations. Therefore the semantics of the dynamic language does not
have to be replicated in a JIT. We call this approach \emph{meta-tracing}.
Another commonality of these approaches is that they allow some annotations (or
hints) in the dynamic language implementation to guide the meta-tracer. This
makes the process not completely automatic but can give good speedups over
bare meta-tracing.

@@

Concretely these hints are used to control how the optimizer of the
tracing JIT can improve the traces of the object model. More
specifically, these hints influence the constant folding
optimization. The first hint makes it possible to turn arbitrary
variables in the trace into constant by feeding back runtime values. The
second hint allows the definition of additional foldable operations.

@@

A tracing JIT works by recording traces of concrete execution paths through the
program. Those
traces are therefore linear list of operations, which are optimized and then
get turned into machine code. This recording automatically inlines functions:
when a function call is encountered the operations of the called functions are
simply put into the trace too.

@@

To be able to do this recording, VMs with a
tracing JIT typically contain an interpreter. After a user program is
started the interpreter is used; only the most frequently executed paths through the user
program are turned into machine code. The tracing JIT tries to produce traces
that correspond to loops in the traced program, but most tracing JITs now also
have support for tracing non-loops \cite{andreas_gal_incremental_2006}.

One disadvantage of (tracing) JITs which makes them not directly applicable to
PyPy is that they need to encode the language semantics of the language they are
tracing. Since PyPy wants to be a
general framework, we want to reuse our tracer for different languages.
Therefore PyPy's JIT is a meta-tracer \cite{bolz_tracing_2009}. It does not
trace the execution of the user program, but instead traces the execution of
the \emph{interpreter} that is running the program. This means that the traces
it produces don't contain the bytecodes of the language in question, but
RPython-level operations that the interpreter did to execute the program.

@@

While the operations in a trace are those of the interpreter, the loops that are
traced by the tracer are the loops in the
user program. This means that the tracer stops tracing after one iteration of
the loop in the user function that is being considered. At this point, it probably
traced many iterations of the interpreter main loop.

@@

Before sending the trace to the backend to produce actual machine code, it is
optimized.  The optimizer applies a number of techniques to remove or simplify
the operations in the trace. Most of these are well known compiler optimization
techniques, with the difference that it is easier to apply them in a tracing
JIT because it only has to deal with linear traces.  Among the techniques:
%
\begin{itemize}
    \item constant folding
    \item common subexpression elimination
    \item allocation removal \cite{bolz_allocation_2011}
    \item store/load propagation
    \item loop invariant code motion
\end{itemize}

@@

@@

In the trace above, the value of $x_1$ is statically known thanks to the
guard. Remember that a guard is a runtime check. The above trace will run to
completion when $x_1$ \texttt{== 4}. If the check fails, execution of the trace is
stopped and the interpreter continues to run.

There are cases in which it is useful to turn an arbitrary variable
into a constant value. This process is called \emph{promotion} and it is an old idea
in partial evaluation (it's called ``The Trick''  \cite{jones_partial_1993} there). Promotion is also heavily
used by Psyco \cite{rigo_representation-based_2004} and by all older versions
of PyPy's JIT. Promotion is a technique that only works well in JIT compilers;
in static compilers it is significantly less applicable.

Promotion is essentially a tool for trace specialization. There are places in
the interpreter where knowing that a value is constant opens a lot of
optimization opportunities, even though it
could have different values in practice. In such a place, promotion is used. The
typical reason to do that is if there is
a lot of computation depending on the value of that variable.

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$v_1$ = $x_1$ * 2
$z_1$ = $v_1$ + 1
$v_2$ = $z_1$ + $y_1$
return(v2)
\end{lstlisting}

The promotion is turned into a \texttt{guard} operation in the trace. The guard
captures the value of $x_1$ as it was at runtime. From the point of view of the
optimizer, this guard is not any different than the one produced by the \texttt{if}
statement in the example above. After the guard, the rest of the trace can
assume that $x_1$ is equal to \texttt{4}, meaning that the optimizer will turn this
trace into:

@@

This new trace will be attached to the guard instruction of the first trace. If
$x_1$ takes on even more values, a new trace will eventually be made for all of them,
linking them into a chain. This is clearly not desirable, so we should promote
only variables that don't vary much. However, adding a promotion hint will never produce wrong
results. It might just lead to too much assembler code.

Promoting integers, as in the examples above, is not used that often.
However, the internals of dynamic language interpreters often
have values that are variable but vary little in the context of parts of a user
program. An example would be the types of variables in a user function. Even
though in principle the argument to a Python function could be any Python type,
in practice the argument types tend not to vary often. Therefore it is possible to
promote the types. The next section will present a complete example of how
this works.

@@

@@

On the one hand, the \texttt{purefunction} annotation is very powerful. It can be
used to constant-fold arbitrary parts of the computation in the interpreter.
However, the annotation also gives the interpreter author ample opportunity to mess things up. If a
function is annotated to be pure, but is not really, the optimizer can produce
subtly wrong code. Therefore, a lot of care has to be taken when using this
annotation.

@@

\subsubsection{Immutable Fields}

One of the most common cases of pure functions is reading immutable
values out of objects. Since this is so common, we have special syntactic sugar
for it. A RPython class can have a class attribute \texttt{\_immutable\_fields\_} set to
a list of strings, listing the fields that cannot be changed. This is equivalent
to using getters and annotating them with \texttt{purefunction}.

@@

In this implementation instances no longer use dictionaries to store their fields. Instead, they have a
reference to a map, which maps field names to indexes into a storage list. The
storage list contains the actual field values. The maps are shared between
objects with the same layout. Therefore they have to be immutable, which means
that their \texttt{getindex} method is a pure function. When a new attribute is added
to an instance, a new map needs to be chosen, which is done with the
\texttt{add\_attribute} method on the previous map (which is also pure). Now that we have
introduced maps, it is safe to promote the map everywhere, because we assume
that the number of different instance layouts is small.

@@

Therefore, we give every class a version object, which is changed every time a
class gets changed (i.e., the content of the \texttt{methods} dictionary changes).
This means that the result of \texttt{methods.get()} for a given \texttt{(name,
version)} pair will always be the same, i.e. it is a pure operation.  To help
the JIT to detect this case, we factor it out in a helper method which is
explicitly marked as \texttt{@purefunction}. The refactored \texttt{Class} can
be seen in Figure~\ref{fig:version}

@@

For space reasons we cannot perform a full evaluation here, but still want to
present some benchmark numbers. We chose to present just some benchmarks: The
templating engine of the Django web
framework\footnote{\texttt{http://www.djangoproject.com/}}; a Monte-Carlo Go
AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/
disco-elegant-python-go-player.html}}; a BZ2 decoder; a port of the classical
Richards benchmark in Python; a Python version of the Telco decimal
benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}}, using a
pure Python decimal floating point implementation. The results we see in these
benchmarks seem to repeat themselves in other benchmarks using object-oriented
code; for purely numerical algorithms the speedups are significantly smaller.

The benchmarks were run on an otherwise idle Intel Core2 Duo P8400 processor
with 2.26 GHz and 3072 KB of cache on a machine with 3GB RAM running Linux
2.6.35. We compared the performance of various Python implementations on the
benchmarks. As a baseline, we used the standard Python implementation in C,
CPython 2.6.6\footnote{\texttt{http://python.org}}, which uses a bytecode-based
interpreter. We compare it against four versions of PyPy's Python interpreter,
all of them with JIT enabled. The PyPy baseline does not enable maps or type
versions. We then benchmarked PyPy, first using each technique separately,
and finally using both together.

All benchmarks were run 50 times in the same process, to give the JIT time to
produce machine code. The arithmetic mean of the times of the last 30 runs were
used as the result. The errors were computed using a confidence interval with a
95\% confidence level \cite{georges_statistically_2007}. The results are
reported in Figure~\ref{fig:times}, together with the same numbers normed to
those of the full JIT.

>>>>

It has long been an objective of the partial evaluation community to
automatically produce compilers from interpreters. There has been a
renaissance of this idea around the approach of tracing just-in-time
compilers. A number of projects have attempted this approach. SPUR \cite{bebenita_spur:_2010} is
a tracing JIT for .NET together with a JavaScript implementation in C\#. PyPy
\cite{armin_rigo_pypys_2006} contains a tracing JIT for RPython \cite{davide_ancona_rpython:_2007} (a restricted
subset of Python). This JIT is then used to trace a number of languages
implementations written in RPython. A number of other experiments in this
directions were done, such as an interpreter for Lua in JavaScript, which is run
on and optimized with a tracing JIT for JavaScript
\cite{yermolovich_optimization_2009}.

These projects have in common that they work one meta-level down, providing a tracing JIT for the
language used to implement the dynamic language, and not for the dynamic language itself.
The tracing JIT will then trace through the object model of the dynamic
language implementation. This makes the object model transparent to the tracer
and its optimizations. Therefore the semantics of the dynamic language does not
have to be replicated in a JIT. We call this approach \emph{meta-tracing}.
Another commonality of these approaches is that they allow some annotations (or
hints) in the dynamic language implementation to guide the meta-tracer. This
makes the process not completely automatic but can give good speedups over
bare meta-tracing.

@@

Concretely these hints are used to control how the optimizer of the
tracing JIT can improve the traces of the object model. In particular the hints
influence the constant folding
optimization. The first hint makes it possible to turn arbitrary
variables in the trace into constant by feeding back runtime values. The
second hint allows the definition of additional foldable operations.

@@

A tracing JIT works by recording traces of concrete execution paths through the
program. Those
traces are therefore linear list of operations, which are optimized and then
get turned into machine code. This recording automatically inlines functions:
when a function call is encountered the operations of the called functions are
simply put into the trace of the caller too. The tracing JIT tries to produce traces
that correspond to loops in the traced program, but most tracing JITs now also
have support for tracing non-loops \cite{andreas_gal_incremental_2006}.

@@

To be able to do this recording, VMs with a
tracing JIT typically contain an interpreter. After a user program is
started the interpreter is used; only the most frequently executed paths through the user
program are turned into machine code. The interpreter is also used when a guard
fails to continue the execution from the failing guard.

One disadvantage of (tracing) JITs which makes them not directly applicable to
PyPy is that they need to encode the language semantics of the language they are
tracing. Since PyPy wants to be a
general framework, we want to reuse our tracer for different languages.
Therefore PyPy's JIT is a \emph{meta-tracer} \cite{bolz_tracing_2009}. It does not
trace the execution of the user program, but instead traces the execution of
the \emph{interpreter} that is running the program. This means that the traces
it produces don't contain the bytecodes of the language in question, but
RPython-level operations that the interpreter did to execute the program.

@@

While the operations in a trace are those of the interpreter, the loops that are
traced by the tracer are the loops in the
user program. To achieve this the tracer stops tracing after one iteration of
the loop in the user function that is being considered. At this point, it probably
traced many iterations of the interpreter main loop.

@@

Before sending the trace to the backend to produce actual machine code, it is
optimized.  The optimizer applies a number of techniques to remove or simplify
the operations in the trace. Most of these are well known compiler optimization
techniques, with the difference that it is easier to apply them in a tracing
JIT because it only has to deal with linear traces.  Among the techniques are
constant folding, common subexpression elimination, allocation removal
\cite{bolz_allocation_2011}, store/load propagation, loop invariant code
motion.

@@

\cfbolz{should we show the code that would create the inst use in tracing?}

@@

In the trace above, the value of $x_1$ is statically known after the guard.
Remember that a guard is a runtime check. The above trace will run to
completion when $x_1$ \texttt{== 4}. If the check fails, execution of the trace is
stopped and the interpreter continues to run.

There are cases in which it is useful to turn an arbitrary variable
into a constant value. This process is called \emph{promotion} and it is an old idea
in partial evaluation (it's called ``The Trick''  \cite{jones_partial_1993} there). Promotion is also heavily
used by Psyco \cite{rigo_representation-based_2004} and by all older versions
of PyPy's JIT. It is a technique that only works well in JIT compilers;
in static compilers it is significantly less applicable.

Promotion is essentially a tool for trace specialization. There are places in
the interpreter where knowing that a value is constant opens a lot of
optimization opportunities, even though it
could have different values in practice. In such a place, promotion can be used. The
typical reason to do that is if there is
a lot of computation depending on the value of that variable.

@@

\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($x_1$ == 4)
$v_1$ = $x_1$ * 2
$z_1$ = $v_1$ + 1
$v_2$ = $z_1$ + $y_1$
return($v_2$)
\end{lstlisting}

The promotion is turned into a \texttt{guard} operation in the trace. The guard
captures the value of $x_1$ as it was during tracing. \cfbolz{drop the word runtime feedback here?}
From the point of view of the
optimizer, this guard is not any different than the one produced by the \texttt{if}
statement in the example above. After the guard, the rest of the trace can
assume that $x_1$ is equal to \texttt{4}, meaning that the optimizer will turn this
trace into:

@@

This new trace will be attached to the guard instruction of the first trace. If
$x_1$ takes on even more values, a new trace will eventually be made for all of them,
linking them into a chain. This is clearly not desirable, so we should promote
only variables that don't vary much. However, adding a promotion hint will never produce wrong
results. It might just lead to too much assembler code being generated.

Promoting integers, as in the examples above, is not used that often.
However, the internals of dynamic language interpreters often
have values that are variable but vary little in the context of parts of a user
program. An example would be the types of variables in a user function. Even
though in principle the argument to a Python function could be any Python type,
in practice the argument types tend not to vary often. Therefore it is possible to
promote the types. Section~\ref{sec:} will present a complete example of how
this works.

@@

\cfbolz{should we mention that pure functions are not actually called by the optimizer, but the values that are seen during tracing are used?}

@@

On the one hand, the \texttt{purefunction} annotation is very powerful. It can be
used to constant-fold arbitrary parts of the computation in the interpreter.
However, the annotation also gives the interpreter author ample opportunity to mess things up. If a
function is annotated to be pure, but is not really, the optimizer can produce
subtly wrong code. Therefore, a lot of care has to be taken when using this
annotation\footnote{The most common use case of the \texttt{purefunction}
annotation is indeed to declare the immutability of fields. Because it is so
common, we have special syntactic sugar for it.}.

@@

@@

In this implementation instances no longer use dictionaries to store their fields. Instead, they have a
reference to a map, which maps field names to indexes into a storage list. The
storage list contains the actual field values. Therefore they have to be immutable, which means
that their \texttt{getindex} method is a pure function. When a new attribute is added
to an instance, a new map needs to be chosen, which is done with the
\texttt{add\_attribute} method on the previous map. This function is also pure,
because it caches all new instances of \texttt{Map} that it creates, to make
sure that objects with the same layout have the same map. Now that we have
introduced maps, it is safe to promote the map everywhere, because we assume
that the number of different instance layouts is small.

@@

Therefore, we give every class a version object, which is changed every time a
class gets changed (i.e., the \texttt{methods} dictionary changes).
This means that the result of \texttt{methods.get()} for a given \texttt{(name,
version)} pair will always be the same, i.e. it is a pure operation.  To help
the JIT to detect this case, we factor it out in a helper method which is
explicitly marked as \texttt{@purefunction}. The refactored \texttt{Class} can
be seen in Figure~\ref{fig:version}

@@

For space reasons we cannot perform a full evaluation here, but still want to
present some benchmark numbers. We chose to present just some benchmarks: The
templating engine of the Django web
framework\footnote{\texttt{http://www.djangoproject.com/}}; a Monte-Carlo Go
AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/
disco-elegant-python-go-player.html}}; a BZ2 decoder; a port of the classical
Richards benchmark in Python; a Python version of the Telco decimal
benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}}, using a
pure Python decimal floating point implementation. The results we see in these
benchmarks seem to repeat themselves in other benchmarks using object-oriented
code; for purely numerical algorithms the speedups introduced by the techniques
in this paper are much smaller because they are already fast.

The benchmarks were run on an otherwise idle Intel Core2 Duo P8400 processor
with 2.26 GHz and 3072 KB of cache on a machine with 3GB RAM running Linux
2.6.35. We compared the performance of two Python implementations on the
benchmarks. As a baseline, we used the standard Python implementation in C,
CPython 2.6.6\footnote{\texttt{http://python.org}}, which uses a bytecode-based
interpreter. We compare it against two versions of PyPy's Python interpreter,
both of them with JIT enabled. The PyPy baseline does not enable maps or type
version, the full JIT enables both.

All benchmarks were run 50 times in the same process, to give the JIT time to
produce machine code. The arithmetic mean of the times of the last 30 runs were
used as the result. The errors were computed using a confidence interval with a
95\% confidence level \cite{georges_statistically_2007}. The results are
reported in Figure~\ref{fig:times}, together with the same numbers normalized to
those of the full JIT.
